---
title: "Part3"
author: "Patty Frontiera"
date: "Spring 2018"
#output: slidy_presentation
output: ioslides_presentation
---


# Spatial Analysis

## Spatial Analysis

Spatial Analysis begins with an exploration of the data. 

1. Mapping to see its location and distribution

2. Asking questions of, or querying, your data. 

3. Cleaning & reshaping the data

4. Applying analysis methods

5. Mapping analysis results

6. Repeat as needed


## Spatial Queries

There are two key types of spatial queries

- spatial measurement queries, 
    - e.g. area, length, distance

- spatial relationship queries, 
    - e.g. what locations in A are also in B.

These types are often combined, e.g.

- How much of region A that is within region B?

## Our Questions

So far we have explored housing values for the city of San Francisco. 

The data set consists of a lot of dissaggregated features represented as points.

In this section we will

- determine the census tract id (GEOID) for each property
- calculate the average property value per neighborhood.
- determine the average property value within walking dist of coit tower


## R Spatial Libraries

Let's load the R libraries we will use

```{r, eval=F}
library(sp)
library(rgdal)
library(rgeos)
library(tmap)
library(RColorBrewer)
library(ggplot2)
library(ggmap)
## leaflet, ggmap, maptools, ??
```

```{r, echo=F}
library(sp)
library(rgdal)
library(rgeos)
library(tmap)
library(RColorBrewer)
library(ggplot2)
library(ggmap)
```

## R Packages for Spatial Analysis & Mapping

**sp**: Classes and methods for spatial data

**rgdal**: for importing, exporting and transforming spatial data

**rgeos**: for spatial operations and queries on geometric objects

**tmap**: for creating interactive web maps

**RColorBrewer**: for selecting predefined color palettes.

**ggmap/ggplot2** for geocoding locations, mapping and plotting 



## SF Properties 2015

Load the data if it is not already loaded

```{r}
# setwd()
sfhomes <- read.csv('data/sf_properties_25ksample.csv')
sfhomes15 <- subset(sfhomes, as.numeric(SalesYear) == 2015)
class(sfhomes15)
sfhomes15_sp <- sfhomes15
coordinates(sfhomes15_sp) <- c('lon','lat') # ORDER MATTERS!!
proj4string(sfhomes15_sp) <- CRS("+init=epsg:4326")
```
 

## SF Census Tracts

Use `readOGR` to load the shapefile `sf_pop_by_tracts`

- it's in the `data` folder

Then take a look at the data.

- What type of `sp` object is it?
- What is its CRS?
- What column contains the population data?

## SF Census Tracts
```{r}
sftracts <- readOGR(dsn="./data", layer="sf_pop_by_tracts")
class(sftracts)
proj4string(sftracts)
summary(sftracts)
head(sftracts@data)
```

## Make a Quick Plot

```{r}
plot(sftracts)  # or qtm(sftracts)
```


## Remove the Farallon Islands tract

You can subset a SPDF just like a DF

Select all features with population > 0

Then replot the map

## Subset

```{r}
sftracts<- subset(sftracts, pop14 > 0)
plot(sftracts)
```

## What Tract is each property in?

We need to **spatially join** the `sftracts` and `sfhomes15_sp` to answer this.


## Spatial Join

A spatial join associates rows of data in one object with rows in another object based on the spatial relationship between the two objects.

A spatial join is based on the comparison of two sets of geometries in the same coordinate space. 

This is called a **spatial overlay**.

 

## Spatial overlay

Spatial overlay operations in R are implemented using the `over` function in the `SP` and `rGEOS` libraries. 

Point-polygon overlay use SP::over

SpatialLines objects, or pairs of SpatialPolygons require package `rgeos`, and use `gIntersects`. 

That's likely more detail than you need right now but the *point* here is that `rgeos` is the go-to library for vector geometry processing in R.

##  `over`

You can interperet `over(x,y)`  as:

- for each feature in X give me information about the first feature in Y at the corresponding location. 

You can interperet `over(x,y, returnList=TRUE)`  as:

- for each feature in X give me information about the all features in Y at the corresponding location. 

See `?over` for details.


## So here goes...

*In what tract is each SF property is located?* 

```{r, eval=F}
homes_with_tracts <- over(sfhomes15_sp, sftracts)
```

## Did it work?

If not, why not?

# Coordinate reference systems (CRS) must be the same!


## CRSs must be the same

The `over` function, like almost all spatial analysis functions, requires that both data sets be spatial objects (they are) with the same coordinate reference system (CRS). Let's investigate

```{r, eval=F}
# What is the CRS of the property data?
proj4string(sfhomes15_sp)

# What is the CRS of the census tracts?
proj4string(sftracts)
```

## Geographic CRS

Both data are in a geographic CRS but they are different 
- WGS84 (sfhomes15_sp) and NAD83 (sftracts)

1. Transform the CRS of `sftracts` to that of `sfhomes15_sp`

2. Then test that they are identical


## Transform the CRS

```{r}
# Transform the CRS for tracts to be the same as that for sfhomes15_sp
sftracts2 <- spTransform(sftracts, CRS(proj4string(sfhomes15_sp)))

# make sure the CRSs are the same
proj4string(sftracts2) == proj4string(sfhomes15_sp) 
```

Now let's try that overlay operation again

## Try 2

```{r}
# Now try the overlay operation again
# Identify the tract for each property in sfhomes15_sp
homes_with_tracts <- over(sfhomes15_sp, sftracts2)
```

## Questions

What is our output? Does it answer our question?

What type of data object did the over function return?

```{r, eval=F}
head(homes_with_tracts) # take a look at the output
class(homes_with_tracts)
nrow(homes_with_tracts)
nrow(sftracts2)
nrow(sfhomes15_sp)
```


## `over` discussion

Our output *homes_with_tracts* is a data frame that contains 
- the id of each property in `sfhomes15_sp`
- all of the columns from `sftracts2@data` including the census tract id (GEOID) 

So we are close to answering our question.

But for the data to be useful we need 
- to link (or join) the GEOID to the `sfhomes15_sp` object

## Keep just the GEOID column

```{r}

homes_with_tracts <- homes_with_tracts[c("GEOID")]
```

## Add the the DF to the SPDF

We can use the base cbind (column bind) command to join the data frame to the SpatialPointsDataFrame.

*The successful use of this function requires the data to be in the same order!*

```{r}

sfhomes15_sp@data <-cbind(sfhomes15_sp@data, homes_with_tracts)
## NOTE - binding to the @data slot!

# Review and note the change
# head(sfhomes15_sp@data)
```

## Check in `tmap`

Map the data in tmap interactive mode 

##

```{r}
tmap_mode("view")
tm_shape(sftracts2) + tm_polygons(col="beige") + tm_shape(sfhomes15_sp) + 
  tm_dots(col="red")
```

## Evaluation

Did the `over` command successfully associate the census tract GEIOD with each property?

If yes, you now could link the property data to census demographic data by GEOID.


## Point-in-Polygon Queries

We just did what's called a type of spatial overlay query called a `Point-in-Polygon` query. 

We asked "In what tract is each property located?". 

## WOW

Data linkage via space!

## The Census Tract Perspective

We now know the tract for each property.

Now let's think about this question from the tract perspective. 

Let's ask the question

- What is the average propety value per tract?



## Non-Spatial Aggregation

Since we joined GEOID to each property we can use the non-spatial `aggregate` function to compute the mean of totvalues for each GEOID.

```{r}

mean_totvalue_by_tract <- aggregate(totvalue ~ GEOID, sfhomes15_sp, mean)

# Take a look
head(mean_totvalue_by_tract)

```

## Rename the column!

So that it is clear that it is the mean for the tract!
```{r}
colnames(mean_totvalue_by_tract) <- c("GEOID","mean_totvalue")
head(mean_totvalue_by_tract)
```
## Map Tracts by Mean Property Value

However, we can't map our data frame of mean_totvalues by GEOID.

We can use `sp::merge` to join the `mean_totvalue_by_tract` DF to the `sftracts2` SPDF.

We should make sure that

- the number of rows in `sftracts2` and `mean_totvalue_by_tract` are the same

- they share a column of common values - GEOID


## Join by Attribute

When we join two data objects based on values in a column it is called a data table `join by attribute`.

The sp:merge makes this syntax simple for `sp` objects with @data slots.

```{r}

sftracts2<- merge(sftracts2, mean_totvalue_by_tract, 
                  by.x="GEOID", by.y="GEOID")
```

IMPORTANT: DO NOT merge the DF to the @data slot! but rather to the SPDF!

```
## Don't do this:
 sftracts2@data <- merge(sftracts2@data, mean_totvalue_by_tract,
              by.x="GEOID", by.y="GEOID")
```
## Take a look

```{r}
head(sftracts2@data)
```


## Challenge

Create an interactive tmap of census tracts colored by mean_totvalue

##
```{r}
tm_shape(sftracts2) + tm_polygons(col="mean_totvalue", style="jenks")
```

## Aggregating Data Spatially

Above we asked "what is the census tract id for each property?"

We then used the non-spatial `aggregate` function to calculate the mean totvalue for each tract.

Finally, we did a spatial `merge` to join this results to the census tracts.

However, we can ask the same from the tract perspective using `sp::aggregate`

## sp::aggregate

Compute mean totvalue by census tract

- what class is the output object?

```{r}

tracts_with_property_count <- aggregate(x = sfhomes15_sp["totvalue"], 
                                        by = sftracts2, FUN = mean)
```

## Examine output

```{r}
class(tracts_with_property_count)
head(tracts_with_property_count@data)
plot(tracts_with_property_count)
```

## Add the GEOID

```{r}
tracts_with_property_count$GEOID <- sftracts2$GEOID
```

## Check it

Did both methods get the same result? 

Check the mean totvalue for the same GEOID in each SPDF

```{r}

tracts_with_property_count[tracts_with_property_count$GEOID == "06075010700", ]$totvalue 

sftracts2[sftracts2$GEOID == "06075010700",]$mean_totvalue

```

# Distance queries

## Distance queries

Many methods of spatial analysis are based on distance queries.

For example, point pattern analysis considers the distance between features to determine whether or not they are clustered.

We can also use distance as a way to select features spatially.

## Distance

Let's compute the mean  property value within 1 km of Coit Tower

First, we need to know where Coit Tower is. How can we find that out?

## Geocoding a place name with ggmap

```{r}
library(ggmap)
library(ggplot2)

coit_tower <- geocode('Coit Tower, San Francisco, CA')
coit_tower_pt <- c(coit_tower$lon, coit_tower$lat) # -122.40582, 37.80239
coit_tower_pt

```



## Selecting by Distance

In order to select properties with 1KM of Coit Tower we
- create a 1KM radius buffer polygon around the Coit Tower point

We then do a point-in-polygonn operation to either count the number of properties within the buffer or compute the mean totvalue.

## rgeos

`rgeos` is another powerful and widely used library for working with geospatial data.

It is the muscle for 
- creating new geometries from exisiting ones
- calculating spatial metrics like area, length, distance
- calculating the spatial relationship between two geometries.

We can use the `rgeos::gBuffer` function to create our buffer polygon

## Can you understand this code?
```{r}

coordinates(coit_tower) <- c("lon","lat")
proj4string(coit_tower)<- CRS("+init=epsg:4326")

coit_tower_utm <- spTransform(coit_tower, CRS("+init=epsg:26910"))

library(rgeos)
coit_km_buffer <- gBuffer(coit_tower_utm, width=1000)

```

## Map the Buffer

```{r}

tm_shape(sfhomes15_sp) + tm_dots(col="blue") +
tm_shape(coit_km_buffer) + tm_borders(col="red", lwd=2) +
tm_shape(coit_tower_utm) + tm_dots(col="red")
```

## Question

Now that we have our buffer polygon, how can we compute the mean totvalue of properties within the buffer?

## Spatially Aggregate by Buffer

```{r, eval=F}
buff_mean <- aggregate(x = sfhomes15_sp["totvalue"], 
                       by = coit_km_buffer, FUN = mean)
```

## CRS....
```{r}

coit_buffer_lonlat <- spTransform(coit_km_buffer, 
                                  CRS(proj4string(sfhomes15_sp)))

buff_mean <- aggregate(x = sfhomes15_sp["totvalue"], 
                       by = coit_buffer_lonlat, FUN = mean)

```

## Check Results

What is the mean property value within 1KM of Coit Tower in 2015?

`View(buff_mean)`

## Questions?


## Summary

That was a whirlwind tour of just some of the methods of spatial analysis.

There was a lot we didn't and can't cover.

Raster data is a another major topic!
- but the `raster` package is the key

## Spring semester classes

- Geog 88: Geography & Data Science
- Geog 187: Geographic Info Analysis
- LA221: Quantitative Methods in Environmental Planning
- CyPlan 204C: GIS for City Planning


## Selected  References

- [Spatial Data in R tutorial](https://cengel.github.io/rspatial)
- [An Introduction to Spatial Data Analysis and Visualisation in R](https://data.cdrc.ac.uk/tutorial/an-introduction-to-spatial-data-analysis-and-visualisation-in-r)
- [NEON Spatial Data tutorials](http://neondataskills.org/tutorial-series/)
- [GIS in R](http://www.nickeubank.com/gis-in-r)
- [Spatial Data Analysis and Modeling in R](http://www.rspatial.org/index.html)
- [Intro to visualizing Spatial Dat in R](https://github.com/Robinlovelace/Creating-maps-in-R)
- [RStudio Leaflet in R tutorial](https://rstudio.github.io/leaflet)
- [Blog on mapping census data in R](http://zevross.com/blog/2015/10/14/manipulating-and-mapping-us-census-data-in-r-using-the-acs-tigris-and-leaflet-packages-3/)
- [Geocomputation in R online book featuring sf Package](http://robinlovelace.net/geocompr/ )
- [CRAN Task View: Analysis of Spatial Data](https://cran.r-project.org/web/views/Spatial.html)
- http://rstudio-pubs-static.s3.amazonaws.com/6577_3b66f8d8f4984fb2807e91224defa854.html


## Output code to script

```{r, eval=F}
library(knitr)
purl("r-geospatial-workshop-f2017-pt2.Rmd", output = "scripts/r-geospatial-workshop-f2017-pt2.r", documentation = 1)
```
