---
title: "Geospatial Data & Maps in R, pt 2"
author: "Patty Frontiera"
date: "February 2018"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part II Overview

Mapping spatial objects with tmap

Taste of Spatial Analysis

## Part II Prep

1. https://github.com/dlab-geo/r-geospatial-workshop

- Click *Clone or Download* and download the zip file
- Upzip the zip file and make a note of the folder in which it is located

2. Open RStudio and start a new script

3. Set your working directory to where you downloaded and unzipped the files

4. Follow along by opening `r-geospatial-workshop-feb2018-pt2.html` in a web browser

5. Make sure required libraries are installed. 

- sp, rgdal, rgeos, tmap, ggplot2, RColorbrewer, classInt



## R Spatial Libraries

Let's load the R libraries we will use

### Load Libraries
```{r, eval=FALSE}
library(sp)     # spatial objects and methods
library(rgdal)  # read/write from file; manage CRSs
library(rgeos)  # geometric operations
library(tmap)   # mapping spatial objects
```

##
```{r, echo=FALSE}
library(sp)     # spatial objects and methods
library(rgdal)  # read and write from file
library(rgeos)  # geometric operations
library(tmap)   # mapping spatial objects
```
 
## Recap

In the last workshop we:

- Loaded geodata from CSV files into data frames with `read.csv`
- Mapped geographic data frames with `ggplot` and `ggmap`
- Promoted data frames to `SpatialPointsDataFrames` with the `sp::coordinates`
- Loaded geodata from shapefiles with `rgdal::readOGR`
- Defined `CRSs` with the function `sp::proj4string` and `sp::CRS`
- Transformed CRSs with the function `sp::spTransform`
- Saved the sp object `sfboundary_lonlat` as a shapefile using `rgdal::writeOGR`

## Load files from last workshop

Below are the files we will work with.

All of these are in the `data` folder.

- sf_properties_25ksample.csv
- sf_boundary.shp 
- sf_highways.shp
- bart.csv

## Set your working directory

Use `setwd` to set your working directory to the location of the tutorial files.

For example:

```{r, eval=FALSE}
setwd("~/Documents/Dlab/workshops/2018/rgeo/r-geospatial-workshop/r-geospatial-workshop")
```


## SF Properties 2015

Read in from CSV file
```{r}
sfhomes <- read.csv('data/sf_properties_25ksample.csv', stringsAsFactors = FALSE)
sfhomes15 <- subset(sfhomes, as.numeric(SalesYear) == 2015)
class(sfhomes15)
```

## SF Properties 2015

Make it spatial
```{r}
sfhomes15_sp <- sfhomes15  # why do we do this?
coordinates(sfhomes15_sp) <- c('lon','lat') 
proj4string(sfhomes15_sp) <- CRS("+init=epsg:4326")
```

## SF Boundary and SF Highways

Read in the `sf_boundary.shp` and `sf_highways.shp` shapefiles

- what function will you use?

## SF Boundary and SF Highways

```{r}
sf_boundary <- readOGR(dsn="data", layer="sf_boundary")

sf_highways <- readOGR(dsn="data", layer="sf_highways")
```

## SF BART Stations

Read in the `bart.csv` file

Promote the data frame to a `SpatialPointsDataFrame`

## SF BART Stations

```{r}

bart<- read.csv("data/bart.csv", stringsAsFactors = FALSE)

coordinates(bart) <- c("X","Y")
```

## Get the CRS of each sp object

Are they all the same? Describe.
```{r}
proj4string(sf_highways)
proj4string(sf_boundary)
proj4string(bart)
proj4string(sfhomes15_sp)
```

## Define the CRS for SF BART stations

`SpatialPointsDataFrame` will not have a defined CRS if they were created from a data frame

We need to define the CRS of the 'bart' object

What do we need to know to do this

Try it for the `bart` layer

## Terminology

In `GIS` a spatial object is called a `layer`

One element in a layer is called a `feature`

A `feature` consists of the `geometry` and the `attribute data` that describe it

## Define CRS

Where is the CRS definition coming from?

```{r}
proj4string(bart) <- CRS("+init=epsg:4326")
```

*How else could we reference the CRS?*


## Transform data to common CRS

In order to perform spatial analysis we need to first convert all data objects to a common CRS.

Which type? Projected or Geographic CRS?


## Geographic vs. Projected CRS

If my goal is to create maps, I may convert all data to a geographic CRS.

- Which one?

If my goal is to do spatial analysis, I will convert to a projected CRS.

- Which one?


## Common CRS Codes

**Geographic CRSs**

* `4326` Geographic, WGS84 (default for lon/lat)

* `4269` Geographic, NAD83 (USA Fed agencies like Census)


**Projected CRSs**

* `5070` USA Contiguous Albers Equal Area Conic

* `3310` CA ALbers Equal Area

* `26910` UTM Zone 10, NAD83 (Northern Cal)

* `3857` Web Mercator (web maps)


## Transform all layers to UTM 10N, NAD83

Use `spTransform` to transform `bart` and `sfhomes15_sp` to `UTM 10N, NAD83`

`sf_highways` and `sf_boundary` already have this CRS

Recall, this transformation is called `projecting` or `reprojecting`

Try it....

## Transform all layers to UTM 10, NAD83

Note the two methods for doing same thing.

```{r}
sfhomes15_utm <- spTransform(sfhomes15_sp, CRS("+init=epsg:26910"))

bart_utm <- spTransform(bart, CRS(proj4string(sf_boundary)))
```


## Check

Do they match?

```{r}
proj4string(bart_utm) 

proj4string(sfhomes15_utm)
```

## Reproject the CRS AGAIN!

```{r}
#sfhomes15_utm <- spTransform(sfhomes15_sp, CRS("+init=epsg:26910"))

sfhomes15_utm <- spTransform(sfhomes15_sp, CRS(proj4string(sf_boundary)))

# check
proj4string(sfhomes15_utm)
proj4string(sf_boundary)
```



## Map all layers

```{r}
plot(sf_boundary)
lines(sf_highways, col='purple', lwd=4)
points(sfhomes15_utm)
plot(bart_utm, col="red", pch=15, add=T)
```

# tmap

## tmap

`tmap` stands for thematic map

It's a powerful toolkit for creating maps with `sp` objects

yet, with less code than the alternatives

Syntax should be familar to ggplot2 users, but simpler

Relatively easy to create & save interactive maps


## tmap

Load the library

```{r}
library(tmap)
```
You may need to install /load dependencies
- ggplot2, RColorbrewer, classInt, leaflet libraries

## Mapping Spatial Objects

```{r, eval=F}
tm_shape(sf_boundary) + 
  tm_polygons(col="beige", border.col="black")
```

Use `tm_shape(<sp_object>)`  to specifiy a sp object

Add `+ tm_<element>(...)` to style the layer symbology

...and other options for creating a publication ready map

## Mapping polygons

```{r }
tm_shape(sf_boundary) + 
  tm_polygons(col="beige", border.col="black")
```


## tmap starting points

?tmap

?tmap_shape

?tmap_element

- ?tm_polygons (tm_fill, tm_borders)
- ?tm_symbols (tm_dots, etc...)

- [tmap in a Nutshell](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-nutshell.html) 


## Mapping Points

```{r}
tm_shape(sfhomes15_utm) + 
  tm_dots(col="red", size=.25)
```

## Mapping Lines
```{r}
tm_shape(sf_highways) + 
  tm_lines(col="black")
```

## Symbolizing Point Colors by Data Values

```{r}
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25)    # column names must be quoted
```

## Layering data

How do you think we would map all sp objects together?


## Combining the layers
```{r, eval=F}
tm_shape(sf_boundary) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(sf_highways) + 
  tm_lines(col="black") +
tm_shape(sfhomes15_utm) + 
  tm_dots(col="totvalue", size=.25) 
 
```
## Combining the layers
```{r, echo=F}
tm_shape(sf_boundary) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(sf_highways) + 
  tm_lines(col="black") +
tm_shape(sfhomes15_utm) + 
  tm_dots(col="totvalue", size=.25) 
 
```


## CRS management in tmap

Below we are mapping `sf_boundary` (UTM10) and `sfhomes15_sp` (WGS84)
```{r, eval=F}
tm_shape(sf_boundary) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25) 
```

## CRS management in tmap

If the CRSs are defined, `tmap` will use that info

If not, `tmap` will assume `WGS84` 

and dynamically `reproject` subsequent layers to match first one added to map
```{r, echo=F}
tm_shape(sf_boundary) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25) 
```

## Making it more Polished

```{r, eval=F}
tm_shape(sf_boundary) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(sfhomes15_utm) + 
  tm_dots(col="totvalue", size=.25, title = "San Francisco Property Values (2015)") + 
tm_layout(inner.margins=c(.05, .2, .15, .05)) # bottom, left, top, right
```

## Making it more Polished

```{r, echo=F}
tm_shape(sf_boundary) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(sfhomes15_utm) + 
  tm_dots(col="totvalue", size=.25, title = "San Francisco Property Values (2015)") + 
tm_layout(inner.margins=c(.05, .2, .15, .05)) # bottom, left, top, right
```

## Interactive mapping

```{r}
tmap_mode("view")
```

## Handy Shortcut
```{r, eval=F}

last_map()  # display the last map 

```

## Handy Shortcut
```{r}

last_map()  # display the last map 

```

## Reset the mode to static plot

```{r}
tmap_mode("plot")
```


## tmap

Questions?

Check out [tmap in a Nutshell](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-nutshell.html) 

# Spatial Analysis

## Spatial Analysis

1. Mapping / plotting to see location and distribution

2. **Asking questions of, or querying, your data. **

3. Cleaning & reshaping the data

4. Applying analysis methods

5. Mapping analysis results

6. Repeat as needed


## Spatial Queries

There are two key types of spatial queries

- **spatial measurement** queries, 
    - e.g. area, length, distance


- **spatial relationship** queries, 
    - e.g. what locations in A are also in B.

These types are often combined, e.g.

- What is the area of region A is within region B?

# Spatial Measurement Queries

## Computing Area

What is the area of SF Boundary?

- Use `rgeos::gArea` to compute
- What are the units?


```{r}
gArea(sf_boundary)
```

## Area in sq km

Use the `rgeos` function `gArea` to compute area of polygon spatial objects.

- Check results against Wikipedia for [SF](https://en.wikipedia.org/wiki/San_Francisco)

```{r}
gArea(sf_boundary) / (1000 * 1000)
```

## Length of highways

Use the `rgeos` function `gLength` to compute length of linear spatial objects.
```{r}
gLength(sf_highways) / 1000
```

## Distance

The `rgeos` function `gDistance` will return the min distance between two geometries.

Distance in KM between Embarcadero & Powell St Bart stations

- You can check on Google Maps (always spot check!)
```{r}
gDistance(bart_utm[bart_utm$STATION == 'EMBARCADERO',],
          bart_utm[bart_utm$STATION == 'POWELL STREET',]) /1000
```

## Distance between locations

Get distance between all `sfhomes` and `Embarcadero` BART

```{r}
dist2emb<- gDistance(bart_utm[bart_utm$STATION == 'EMBARCADERO',],
                     sfhomes15_utm, byid=TRUE) /1000
# check output
nrow(dist2emb)
nrow(sfhomes15_utm)
head(dist2emb)
```

## Distance between locations

*How is the output different if `byid=FALSE`* (default)
```{r}
#dist2emb<- gDistance(bart_utm[bart_utm$STATION == 'EMBARCADERO',],
#                     sfhomes15_utm, byid=TRUE) /1000

gDistance(bart_utm[bart_utm$STATION == 'EMBARCADERO',],
                     sfhomes15_utm) /1000
```


# Spatial Relationships

## Spatial Relationships

**Spatial relationship queries** compare the geometries of two spatial objects in the same coordinate space (CRS).

There are many, often similar, functions to perform these queries (can be confusing!).

These operations may return logical values, lists, matrices, dataframes, geometries or spatial objects

- you need to check what type of object is returned 

- you need to check what values are returned to make sure they make sense

## Spatial Relationship Queries

CRSs `DO` need to match

But ...

they don't need to be Projected CRSs

They can be any CRS


## Are there any BART stations in SF?

This is a very common type of spatial query called a `point-in-polygon` query.

We can use the `rgeos` function `gIntersects` to answer this.

We already know the answer but we want to see how it is done.


## Are there any BART stations in SF?

*What is the output of gIntersects?*

 - uncomment to find out
 
```{r, eval=FALSE}
bart_stations_in_sf <-gIntersects(bart_utm, sf_boundary) 

# bart_stations_in_sf
```

## Are there any BART stations in SF?

What is the output of gIntersects?

- uncomment to find out!
```{r}
bart_stations_in_sf <-gIntersects(bart_utm, sf_boundary)

bart_stations_in_sf
```

## WHAT Bart stations are in SF

Same function, but `byid=TRUE`
What is the output of gIntersects this time?

- uncomment to find out!

```{r}
sfbart_stations <-gIntersects(bart_utm, sf_boundary, byid=TRUE)

# class(sfbart_stations)
# sfbart_stations
```


## What Bart stations are in SF

What is the output of gIntersects?

- why is it different?
```{r}
sfbart_stations <-gIntersects(bart_utm, sf_boundary, byid=TRUE)

class(sfbart_stations)
sfbart_stations
```

*Why is it different?*

## What are the names of the SF BART stations?

We can use the result of `gIntersects` to view the names of the bart stations

```{r}
bart_utm[as.vector(sfbart_stations),]$STATION
```


## Subset the SPDF

We can use the result of `gIntersects` to subset `bart_utm`

```{r}
sfbart_utm <- bart_utm[as.vector(sfbart_stations),]
```



## Map the BART stations

```{r}
tmap_mode("view")

tm_shape(sf_boundary) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(sfbart_utm) + 
  tm_dots(col="red")
```



## Reset `tmap` to plot mode


```{r}
tmap_mode("plot")
```

## gIntersects recap

`rgeos::gIntersects(geom1, geom2)` returns 

- returns a single TRUE/FALSE value indicating if any feature in geom1 intersects any feature in geom2 
    - if `byid=FALSE` (the default when not specified)


- returns a TRUE/FALSE matrix indicating what features in geom1 intersect features geom2 
    - if `byid=TRUE` 


## SF Census Tracts

Let's consider the `sfhomes15_utm` data along with SF Census tract data 


## SF Census Tracts

Use `readOGR` to load the shapefile `sftracts_wpop.shp`

- it's in the `data` folder

Explore the data

- What type of `sp` object is it?
- What is its CRS?
- What column contains the population data?

## SF Census Tracts
```{r}
sftracts <- readOGR(dsn="./data", layer="sftracts_wpop")

class(sftracts)
proj4string(sftracts)

```

## SF Census Tracts
```{r}
head(sftracts@data)
```

## Make a Quick Plot

```{r}
plot(sftracts)  
```

## Spatial Join

A spatial join associates rows of data in one object with rows in another object based on the spatial relationship between the two objects.

A spatial join is based on the comparison of two sets of geometries in the same coordinate space. 

 - This is also called a **spatial overlay**.

The default spatial relationship is `intersects`




## In what census tract is each property located?

We need to **spatially join** the `sftracts` and `sfhomes15_utm` to answer this.

What spatial object are we joining data from? to?

 
## sp::over / rgeos::over

Spatial overlay operations in R are implemented using `over`

Simple point-in-polygon operations use `sp::over`

More complex geometric comparisons use `rgeos::over`

You need to have `rgeos` package installed!

##  `over(x,y)`

You can interperet `over(x,y)`  as:

- for each feature in X give me information about the first feature in Y at the corresponding location. 

You can interperet `over(x,y, returnList=TRUE)`  as:

- for each feature in X give me information about the all features in Y at the corresponding location. 

See `?over` for details.


## So here goes...

*In what census tract is each SF property located?* 

```{r, eval=F}
homes_with_tracts <- over(sfhomes15_utm, sftracts)
```

## Did it work?

If not, why not?


# Coordinate reference systems (CRS) must be the same!


## CRSs must be the same

The `over` function, like almost all spatial analysis functions, requires that both data sets be spatial objects (they are) with the same coordinate reference system (CRS). Let's investigate

```{r, eval=F}
# What is the CRS of the property data?
proj4string(sfhomes15_utm)

# What is the CRS of the census tracts?
proj4string(sftracts)
```


## Transform the CRS

```{r}
# Transform the CRS for tracts to be the same as that for sfhomes15_sp
sftracts_utm <- spTransform(sftracts, CRS(proj4string(sfhomes15_utm)))

# make sure the CRSs are the same
proj4string(sftracts_utm) == proj4string(sfhomes15_utm) 
```

Now let's try that overlay operation again

## Try 2

*In what tract is each SF property is located?* 

```{r}

homes_with_tracts <- over(sfhomes15_utm, sftracts_utm)
```




## `over` output

What is our output? Does it answer our question?

What type of data object did the over function return?

Do we need to add `returnList = TRUE`

```{r, eval=F}
homes_with_tracts <- over(sfhomes15_utm, sftracts_utm)

class(homes_with_tracts)
nrow(homes_with_tracts)
nrow(sftracts_utm)
nrow(sfhomes15_utm)
```

## Review the Over output

```{r}
homes_with_tracts <- over(sfhomes15_utm, sftracts_utm)
class(homes_with_tracts)
nrow(homes_with_tracts)
nrow(sftracts_utm)
nrow(sfhomes15_utm)

```

## Review the over output
```{r}
head(homes_with_tracts)
```



## `over` discussion

Our output *homes_with_tracts* is a **data frame** that contains 

- the same number of rows as `sfhomes15_utm` - in same order
- the id of each property in `sfhomes15_utm`
- the data columns from `sftracts_utm@data` including the census tract id (GEOID) 

So we are close to answering our question.

But for the data to be useful we need 
- to add the `GEOID` column in `homes_with_tracts` to `sfhomes15_utm`

## Add the GEOID column

*CAUTION: this only works because the data are in the same order!*
```{r}
sfhomes15_utm$home_geoid <- homes_with_tracts$GEOID

# Review and note the change
#head(sfhomes15_utm@data)
```


## WOW

Data linkage via space!

The `over` operation gave us the census tract data info for each point in `sfhomes15_utm`

We added the `GEOID` for each point to the `@data` slot of `sfhomes15_utm`

We can now join `sfhomes15_utm` points by `GEOID` to any census variable, eg median household income, and then do an analysis of the relationship between, for example, property value and that variable.

**How would we do that?**

## Read in the census data

Because `GEOIDs` can have leading zeros, we set the `colClasses` to make sure they are not stripped.
```{r}
med_hh_inc <- read.csv("data/sf_med_hh_income2015.csv", stringsAsFactors = F, 
                       colClasses = c("character","numeric"))

head(med_hh_inc)
```

## Spatial*DataFrame to Data Frame Joins

We can use `sp::merge` to join the `med_hh_inc` DF to the `sfhomes15_utm` SPDF.

We should make sure that they share a column of common values - GEOID / home_geoid

We use `sp:merge` not regular `merge` to maintain the integrity of the `sp` object

  - i.e., make sure that the geometry and attribute data slots all stay aligned.

## Join by Attribute

Join two data objects based on common values in a column.

Use `sp:merge` to join a data frame to a spatial object with a data frames (`S*DF` objects)

```{r}

sfhomes15_utm <- merge(sfhomes15_utm, 
                       med_hh_inc, by.x="home_geoid", by.y="GEOID")
```

**IMPORTANT:** DO NOT merge a data frame to the @data slot! 

-Join it to the SPDF!

## Take a look at output

```{r}
head(sfhomes15_utm@data, 2)  # take a look 
```


## Check the `sp::merge` results

```{r}
tmap_mode("view")
tm_shape(sfhomes15_utm) + tm_dots(col="medhhinc")
```

## Reset plot mode
```{r}
tmap_mode("plot")
```

## The Census Tract Perspective

We now know the tract for each property.

Now let's think about this question from the tract perspective. 

Let's ask the question

- What is the average propety value per tract?



## Non-Spatial Aggregation

Since we joined GEOID to each property we can use the non-spatial `aggregate` function to compute the mean of totvalues for each GEOID.

But let's use the `sp` aggregate function instead!

It's actually more straight forward.

## sp::aggregate

Aggregate data values in one spatial object by the geometry of another using the specified function.

```{r}

tracts_with_mean_val <- aggregate(x = sfhomes15_utm["totvalue"], 
                                        by = sftracts_utm, FUN = mean)
```

Wow, so simple. What does that give us?


## Examine output

What type of spatial object is returned by `sp::aggregate`

What values are in the `@data` slot?
```{r}
class(tracts_with_mean_val)
head(tracts_with_mean_val@data)
nrow(tracts_with_mean_val) == nrow(sftracts_utm)
```

## sp::aggregate output

`sp::aggregate` returned a SpatialPolygonsDataFrame

The SPDF has the same geometry as `sftracts_utm`

But the `tracts_with_mean_val@data` slot only contains the mean totvalue for each tract.

To make these data more useful, let's add this value to `sftracts_utm`

## 

This only works because their are the same number of elements both @data slots and they are in the same order!
```{r}

sftracts_utm$mean_totvalue <- tracts_with_mean_val$totvalue

head(sftracts_utm@data) # check it
```

## Map it - do the results make sense

```{r}
tmap_mode("view")
tm_shape(sftracts_utm) +
  tm_polygons(col="mean_totvalue", border.col=NA)
```

## Why no values for some tracts?
```{r}
tmap_mode("view")
tm_shape(sftracts_utm) +
  tm_polygons(col="mean_totvalue", border.col=NA) +
tm_shape(sfhomes15_utm) + tm_dots()
```

# Distance

## Distance queries

Many methods of spatial analysis are based on distance queries.

For example, point pattern analysis considers the distance between features to determine whether or not they are clustered.

We can also use distance as a way to select features spatially.

## Proximity

The other side of distance!

What properties are within walking distance of BART?

## Selecting by Distance

In order to select properties with 1KM of BART
- create a 1KM radius buffer polygon around each BART point

We then do a point-in-polygon operation to either count the number of properties within the buffer or compute the mean totvalue.

## rgeos

`rgeos` is the muscle for 

- creating new geometries from exisiting ones
- calculating spatial metrics like area, length, distance
- calculating the spatial relationship between two geometries.

We can use the `rgeos::gBuffer` function to create our buffer polygon

## Creating Buffers

The `rgeos::gBuffer` function takes as input a spatial object or objects to buffer and a buffer distance.

Let's assume 1KM is standard walking distance.
```{r}

bart_1km_buffer <- gBuffer(sfbart_utm, width=1000)

```

## Map the Buffer
```{r}
tmap_mode("view") 
tm_shape(bart_1km_buffer) + tm_polygons(col="red") +
tm_shape(sfbart_utm) + tm_dots()
```

## Try This

How does this differ when we add `byid=TRUE`
```{r}
bart_1km_buffer_byid <- gBuffer(sfbart_utm, width=1000, byid=TRUE)

tmap_mode("view")
tm_shape(bart_1km_buffer_byid) + tm_polygons(col="red") +
tm_shape(sfbart_utm) + tm_dots()


```

## What sfhomes are within 1km of any bart station?

What operation could we use?

Which buffer polygons?

##  What sfhomes are within 1km of any bart station?

Why `byid=TRUE`

What is the output?
```{r}
homes_near_bart <-  gIntersects(bart_1km_buffer, sfhomes15_utm, byid=TRUE) 
class(homes_near_bart)
head(homes_near_bart)
```

## Select the sfhomes15_utm points in the buffer

```{r}
# subset
sfhomes15_utm_near_bart <- sfhomes15_utm[as.vector(homes_near_bart),]
```

## Map it
```{r}
tm_shape(bart_1km_buffer) + tm_polygons(col="red") +
tm_shape(sfhomes15_utm_near_bart) + tm_dots()
```


## Questions


## Summary

That was a whirlwind tour of just some of the methods of spatial analysis.

There was a lot we didn't and can't cover.

`sf` package is emerging and will eclipse `sp`

Raster data is a another major topic!
- but the `raster` package is the key


## Selected  References & Tutorials

- [Spatial Data in R tutorial](https://cengel.github.io/rspatial)
- [An Introduction to Spatial Data Analysis and Visualisation in R](https://data.cdrc.ac.uk/tutorial/an-introduction-to-spatial-data-analysis-and-visualisation-in-r)
- [NEON Spatial Data tutorials](http://neondataskills.org/tutorial-series/)
- [GIS in R](http://www.nickeubank.com/gis-in-r)
- [Spatial Data Analysis and Modeling in R](http://www.rspatial.org/index.html)
- [Intro to visualizing Spatial Dat in R](https://github.com/Robinlovelace/Creating-maps-in-R)
- [RStudio Leaflet in R tutorial](https://rstudio.github.io/leaflet)
- [Blog on mapping census data in R](http://zevross.com/blog/2015/10/14/manipulating-and-mapping-us-census-data-in-r-using-the-acs-tigris-and-leaflet-packages-3/)
- [Geocomputation in R online book featuring sf Package](http://robinlovelace.net/geocompr/ )
- [CRAN Task View: Analysis of Spatial Data](https://cran.r-project.org/web/views/Spatial.html)
- http://rstudio-pubs-static.s3.amazonaws.com/6577_3b66f8d8f4984fb2807e91224defa854.html


## Output code to script

```{r, eval=F}
library(knitr)
purl("r-geospatial-workshop-feb2018-pt2.Rmd", output = "scripts/r-geospatial-workshop-feb2018-pt2.r", documentation = 2)
```


